---
title: "Data_exploration"
author: "Jessica McCordic"
date: "2026-08-05"
output: html_document
---


# Sequence analysis 

#### Borrowing heavily from this tutorial: https://quantdev.ssri.psu.edu/sites/qdev/files/sequence_tutorial_2017_1122.html



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)



# Load libraries

library(tidyverse)
library(cluster)
library(dendextend)
library(factoextra)
library(fpc)
library(psych)
library(TraMineR)
library(TraMineRextras)


```

```{r load data, echo=FALSE, include=FALSE}

# Load selection tables

# Amplitude metrics (BB SPL, OL, TOL) at 1-second resolution with vessel behavior and GPS
metrics_og <- read_csv("data/Sound_metrics/OLE_202209_coords_metrics.csv")


# fine scale selections of each change in amplitude within vessel passages
selns_int_og <- read_delim("data/Raven_selections/6125.220928161024.Table.1.selections_intervals_JM.txt")


```

### Explore data
```{r explore data}

head(selns_int_og)

plot(selns_int_og$Selection, selns_int_og$`Dur 90% (s)`)
hist(selns_int_og$`Dur 90% (s)`, breaks = 100)

```

### Merge and reshape tables
For sequence analysis, we want (1) a wide df with one row per sequence ID with events (A/B) going across the columns
and (2) a df with sequence ID and category (fishing/transit)

Sequence ID = each line & behavior combo
Event = 2 states (A, B -- alternating per selection)
Class = 2 states (fishing, transit -- determined during field trial)

```{r reshape}

# first we need all the data together to map the categories onto the sub-selections

selns_data <- selns_int_og |>
  mutate(datetime = as_datetime(`Begin Date Time`),
         # assign alternating event codes
         event_ab = rep_len(c("a","b"), length.out = n()))

# map categories from metrics df onto selections
intervals_cat <- selns_data |>
  # join dfs together
  full_join(metrics_og) |>
  # sort by datetime
  arrange(datetime)|>
  # fill in data from metrics df into selection rows
  # fill selection number down to assign metrics rows to selections
  fill(c(Selection,event_ab, Line_num, activity,lat,lon, GPS_speed_kt, Dist_slant))|>
  # create sequence identifier of line + activity
  mutate(seq_ID = 
           case_when(activity == "Off_eff_transit" ~ activity,
                     TRUE ~ paste(Line_num, activity, sep = "_"))) |>
  # filter to just get times included in selection table
  filter(seq_ID != "NA_NA") 
### Want to do something like... 
# when selection changes within seq ID, change event label
# so would end up with 
# seq_ID  | Selection   | Event
# 1           1             A
# 1           1             A
# 1           2             B
# 1           2             B
# 1           2             B
# 1           3             A
# 2           4             A
# 2           4             A
# 2           4             A
# 2           5             B

# etc.


# mutate(event_ab = rep(c("a","b"), each = n())
#   )



selns_seq <- intervals_cat |>
  # get rid of rows without selection info
  filter(!is.na(Selection),
         seq_ID != "NA_NA") |>
 # only keep columns needed for sequence analysis
  select(seq_ID,
         event_ab) |>
 # create time index variable 
  group_by(seq_ID) |>
  mutate(time_idx = seq(1:n())) 


# create key for sequence IDs to fishing vs. transit
fish_states <- selns_seq |>
  select(seq_ID) |>
  mutate(fish_state = case_when(grepl("dredge", seq_ID)==TRUE ~ "fishing",
                   TRUE ~ "transit"))

```

### Descriptives

How many sequences?
How many events per sequence?
How many events of each type across the dataset?

```{r describe dataset}

length(unique(selns_seq$seq_ID))
#  [1] 20
length(unique(fish_states$seq_ID))
#  [1] 20
```

There are 20 sequences in both datasets. 

```{r events}

event_summ <- selns_seq |>
  group_by(seq_ID) |>
  summarize(n_events = n()) |>
  summarize(mean_n = mean(n_events),
            med_n = median(n_events),
            min_n = min(n_events),
            max_n = max(n_events))

event_summ

behav_table <- table(selns_seq$event_ab)

```

### Creating sequences

```{r reformat}

# reformat from long to wide

selns_seq_wide <- selns_seq |>
  filter(seq_ID != "Off_eff_transit") |>
  pivot_wider(names_from = time_idx,
              values_from = event_ab)


```

Create alphabet and assign labels

```{r alphabet}

# this contains categories that appear in the dataset
event_alphabet <- c("a","b")

# formally assigning labels which might be helpful later
event_labels <- c("a", "b")

# assign colors (this feels unnecessary but sure)

event_colors <- viridis::rocket(n = 2) 


```

Create sequence object 

```{r sequence}

# define sequence using seqdef from TraMineR package
event_seq <- seqdef(selns_seq_wide, 
                    var = 2:82,
                    alphabet = event_alphabet,
                    labels = event_labels,
                    xtstep = 5,
                    cpal = event_colors)

# plot sequences

seqIplot(event_seq)


```


Cost matrix

- Use standard cost matrix values:
insertion/deletion cost = 1.0
missingness cost = 0.5*[highest cost in matrix]
substitution cost = constant


```{r cost and dist}

costmatrix <- seqsubm(event_seq, 
                      method="CONSTANT",
                      cval = 2, 
                      with.missing=TRUE,
                      miss.cost=1, 
                      time.varying=FALSE, 
                      weighted=TRUE,
                      transition="both", 
                      lag=1)

costmatrix


# optimal matching dissimilarity matrix

dist_om <- seqdist(event_seq, 
                   method = "OM", 
                   indel = 1.0, # insertion/deletion cost = 1
                   sm = costmatrix, # sub matrix = cost matrix
                   with.missing = TRUE)


# print out top left corner of dissimilarity matrix
# sanity check that it's symmetrical with diagonal all 0's
dist_om[1:10, 1:10]

```


### Clustering

```{r cluster}

event_clusters <- agnes(dist_om, diss = TRUE, method = "ward")
plot(event_clusters, which.plot = 2)

# oof. let's proceed anyways....

event_clust_cut <- cutree(event_clusters, k = 2) |> 
  factor(labels = paste("Group", 1:2))

# plot

seqplot(event_seq, 
        group = event_clust_cut, type = "I")



```


### Group differences between clusters

Visualize dendrogram with cluster ID
Run logistic regression to test whether vessel behavior  1/0 (fishing/transit) differs by cluster membership

```{r test clusters}

# since we're only testing 2 clusters, use logistic regression

# does vessel behavior differ by cluster membership?
# cluster ID = predictor
# vessel behav = response, coded as 1/0 (fishing/transit)

# add cluster labels to seq ID to associate w/fishing vs. transit
seq_clust_lab <- selns_seq_wide |>
  select(seq_ID) |>
  add_column(clust_ID = event_clust_cut)|>
  mutate(fish_state_binom = case_when(grepl("dredge", seq_ID)==TRUE ~ 1,
                   TRUE ~ 0),
         leaf_label = paste(seq_ID, fish_state_binom))

event_clusters$order.lab <- seq_clust_lab$leaf_label[event_clusters$order]
clust_colors = viridis::mako(n = 2, end = 0.75)

fviz_dend(event_clusters, 
          k = 2,
          color_labels_by_k = TRUE,
          k_colors = clust_colors,
          labels_font = "bold",
        
          rect = TRUE,
          rect_border = clust_colors,
          rect_fill = TRUE,
          
          lwd=1,
          horiz = TRUE,
          ggtheme = theme_bw()) +
  theme(axis.text = element_text(size = 10),
        axis.title.x = element_text(size = 12),
        title = element_blank())
```

Some observations from initial dendrogram:

  - Outlier is line 4 towing
  - Lower and Raise mostly clustered with each other with a couple exceptions
  - line transects all mixed in with fishing, no real patterns
  - distant cluster not really diagnostic either


Might as well test it anyway
```{r log reg}
# Logsitic regression
log_reg <- glm(fish_state_binom ~ clust_ID, data = seq_clust_lab, family = binomial())
summary(log_reg)

```

No big surprises. 

# Attempt 2:
### Broader categories, specific events

Want to try more general categories vs. separating out lower/tow/raise

  - seq_ID = Line1_fishing, Line1_transit, Line2_fishing, etc.

Also want to try specifically assigning 'high' and 'low' events
  - calculate mean SPL for whole sequence
  - calculate mean-adjusted SPL (SPL - mean)
  - assign high/low as +/- difference


Re-label dataset
```{r recoding data}


# since we actually care about the levels this time, load in CSV with 1-second measurements

all_metrics <- read_csv("data/Sound_metrics/all.csv")

# merge with selections



hilo_cat <- intervals_cat |>
  # filter out off effort for now
  filter(activity != "Off_eff_transit") |>
  # assign new sequence IDs by fishing activity
  mutate(seq_ID_fish = case_when(grepl("dredge", seq_ID) ==TRUE ~ paste0(as.character(Line_num),"_","fishing"),
                                 TRUE ~ paste0(as.character(Line_num),"_","transit"))
  )|>
  # summarize by sequence to get median metrics across whole sequence
  group_by(seq_ID_fish) |>
  mutate(med_BB = median(BB, na.rm = TRUE),
         med_tol2k = median(TOL_2000, na.rm = TRUE)) |>
  # grop by selection to get median metrics per selection
  group_by(Selection) |>
  mutate(med_sel_BB = median(BB, na.rm = TRUE),
         med_sel_tol2k = median(TOL_2000, na.rm = TRUE),
         diff_BB = med_sel_BB - med_BB,
         diff_tol2k = med_sel_tol2k - med_tol2k, 
         event_hilo = case_when(
           diff_tol2k > 0 ~ "high",
           diff_tol2k < 0 ~ "low")
         )



hilo_seq <- hilo_cat |>
  # get rid of rows without metrics info
  filter(!is.na(event_hilo)) |>
 # only keep columns needed for sequence analysis
  select(seq_ID,
         event_ab) |>
 # create time index variable 
  group_by(seq_ID) |>
  mutate(time_idx = seq(1:n())) 



```










