---
title: "Sequence Analysis"
author: "Jessica McCordic"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---


# Sequence analysis 

#### Borrowing heavily from this tutorial: https://quantdev.ssri.psu.edu/sites/qdev/files/sequence_tutorial_2017_1122.html



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)



# Load libraries

library(tidyverse)
library(cluster)
library(dendextend)
library(factoextra)
library(fpc)
library(psych)
library(TraMineR)
library(TraMineRextras)


```

```{r load data, echo=FALSE, include=FALSE}

# Load selection tables



# Amplitude metrics (BB SPL, OL, TOL) at 1-second resolution
all_metrics_og <- read_csv("data/Sound_metrics/NEFSC_SBNMS_202209_OLE01_allmetrics.csv")

# Amplitude metrics (BB SPL, OL, TOL) measured at 1-second resolution with vessel behavior and GPS
metrics_og <- read_csv("data/Sound_metrics/OLE_202209_coords_metrics.csv")

# fine scale selections of each change in amplitude within vessel passages
selns_int_og <- read_delim("data/Raven_selections/6125.220928161024.Table.1.selections_intervals_JM.txt")


```


### Merge and reshape tables
For sequence analysis, we want (1) a wide df with one row per sequence ID with events (A/B) going across the columns
and (2) a df with sequence ID and category (fishing/transit)

Sequence ID = each line & behavior combo
Event = 2 states (A, B -- alternating by selection)
Class = 2 states (fishing, transit -- determined during field trial)

```{r reshape}

# first we need all the data together to map the categories onto the sub-selections

selns_data <- selns_int_og |>
  mutate(datetime = as_datetime(`Begin Date Time`),
         # assign alternating event codes
         event_ab = rep_len(c("a","b"), length.out = n()))

# pull behaviors & GPS from metrics_og
behav_gps <- metrics_og |>
  select(datetime,Line_num, activity,lat,lon, GPS_speed_kt, Dist_slant)

metrics_join <- all_metrics_og |>
  rename("datetime" = "RLDateTimeHyd") |>
  full_join(behav_gps, by = "datetime", keep = FALSE)


# map categories from metrics df onto selections
intervals_cat <- selns_data |>
  # join dfs together
  full_join(metrics_join) |>
  # get rid of some unused columns
  select(-c(View, Channel,`Low Freq (Hz)`, `High Freq (Hz)` ,`Begin Time (s)`, `End Time (s)`, `Begin Date Time`, OL_Hz:OL_16000, TOL_Hz, TOL_12500:TOL_20000))|>
  # sort by datetime
  arrange(datetime)|>
  
  # fill in data from metrics df into selection rows
  # fill selection number down to assign metrics rows to selections
  fill(c(Selection,event_ab, `Dur 90% (s)`,Line_num, activity,lat,lon, GPS_speed_kt, Dist_slant))|>
  # create sequence identifier of line + activity
  mutate(seq_ID = case_when(activity == "Off_eff_transit" ~ paste(Line_num, activity, sep = "_"),
                            grepl("dredge", activity)==TRUE ~ paste(Line_num, "fishing", sep = "_"),
                            TRUE ~ paste(Line_num, "transit", sep = "_"))) |>
  # filter to just get times included in selection table
  filter(seq_ID != "NA_transit")|>
  group_by(Selection) |>
  mutate(sel_dist = median(round(Dist_slant,0)), 
         sel_dist_cat = case_when(round(sel_dist,0) <= 250 ~ "0-250m",
                                  round(sel_dist,0) > 250 & round(sel_dist,0) <=500 ~ "250-500m",
                                  round(sel_dist,0) > 500 & round(sel_dist,0) <=1000 ~ "500-1000m",
                                  round(sel_dist,0) > 1000 ~ ">1000m"))




# Assign sequence events based on relative SPL (high vs low) and Duration

# get 50th percentile to use as "short" vs "long" cutoff
dur_cut_val <- quantile(unique(intervals_cat$`Dur 90% (s)`))[3]

# start with intervals_cat
hilo_cat <- intervals_cat |>
  # group by seq_ID & selection to get median metrics per selection
  group_by(seq_ID, Selection) |>
  summarize(med_sel_BB = median(BB, na.rm = TRUE),
            med_sel_tol63 = median(TOL_63, na.rm = TRUE),
            med_sel_tol125 = median(TOL_125, na.rm = TRUE),
            med_sel_tol2k = median(TOL_2000, na.rm = TRUE),
            Dur_90_perc = mean(`Dur 90% (s)`)) |>
  
  # assign categories based on SPL metrics and duration
  mutate(
    n_selns = n(), 
    
    ####
    # Broadband SPL
    # get lag and lead values for differences
    lag_BB = lag(med_sel_BB),
    lead_BB = lead(med_sel_BB),
    diff_BB = case_when(
      # if lagged value is NA (AKA first value in group), use lead value to fill it in
      is.na(lag_BB) ~ med_sel_BB - lead_BB,
      
      TRUE ~ med_sel_BB - lag_BB),
    # assign high/low categories based on TOL values per selection within each sequence
    BB_hilo = case_when(
      # For single instance of one selection for whole sequence (Sel 25, Line 4, transit),
      #   compare to previous selection (Sel 24) which is lower, so assign manually as "high"
      n_selns == 1 ~ "high",
      diff_BB > 0 ~ "high",
      diff_BB < 0 ~ "low"
    ),
    
    
    ####
    # TOL63
    # get lag and lead values for differences
    lag_tol63 = lag(med_sel_tol63),
    lead_tol63 = lead(med_sel_tol63),
    diff_tol63 = case_when(
      # if lagged value is NA (AKA first value in group), use lead value to fill it in
      is.na(lag_tol63) ~ med_sel_tol63 - lead_tol63,
      
      TRUE ~ med_sel_tol63 - lag_tol63),
    # assign high/low categories based on TOL values per selection within each sequence
    tol63_hilo = case_when(
      # For single instance of one selection for whole sequence (Sel 25, Line 4, transit),
      #   compare to previous selection (Sel 24) which is lower, so assign manually as "high"
      n_selns == 1 ~ "high",
      diff_tol63 > 0 ~ "high",
      diff_tol63 < 0 ~ "low"
    ),
    
    ####
    # TOL125
    # get lag and lead values for differences
    lag_tol125 = lag(med_sel_tol125),
    lead_tol125 = lead(med_sel_tol125),
    diff_tol125 = case_when(
      # if lagged value is NA (AKA first value in group), use lead value to fill it in
      is.na(lag_tol125) ~ med_sel_tol125 - lead_tol125,
      
      TRUE ~ med_sel_tol125 - lag_tol125),
    # assign high/low categories based on TOL values per selection within each sequence
    tol125_hilo = case_when(
      # For single instance of one selection for whole sequence (Sel 25, Line 4, transit),
      #   compare to previous selection (Sel 24) which is lower, so assign manually as "high"
      n_selns == 1 ~ "high",
      diff_tol125 > 0 ~ "high",
      diff_tol125 < 0 ~ "low"
    ),
    
    
    # TOL2000
    # get lag and lead values for differences
    lag_tol2k = lag(med_sel_tol2k),
    lead_tol2k = lead(med_sel_tol2k),
    diff_tol2k = case_when(
      # if lagged value is NA (AKA first value in group), use lead value to fill it in
      is.na(lag_tol2k) ~ med_sel_tol2k - lead_tol2k,
      
      TRUE ~ med_sel_tol2k - lag_tol2k),
    # assign high/low categories based on TOL values per selection within each sequence
    tol2k_hilo = case_when(
      # For single instance of one selection for whole sequence (Sel 25, Line 4, transit),
      #   compare to previous selection (Sel 24) which is lower, so assign manually as "high"
      n_selns == 1 ~ "high",
      diff_tol2k > 0 ~ "high",
      diff_tol2k < 0 ~ "low"
    ),
    
    # assign long/short categories based on 90% duration per selection within each sequence
    dur_long_short = case_when(
      Dur_90_perc <= dur_cut_val ~ "short",
      Dur_90_perc > dur_cut_val ~ "long"
    ),
    
    # paste SPL and duration combos together (4 possible values)
    event_hilo_tol63 = paste0(tol63_hilo, "_", dur_long_short),
    event_hilo_tol125 = paste0(tol125_hilo, "_", dur_long_short),
    event_hilo_tol2k = paste0(tol2k_hilo, "_", dur_long_short),
    event_hilo_BB = paste0(BB_hilo, "_", dur_long_short))

#### BB SPL
# set up df for sequence analysis  
selns_seq_BB <- hilo_cat |>
  # filter out off effort for now
  filter(seq_ID != "Off_eff_transit") |>
  ungroup()|>
  group_by(seq_ID) |>
  # only keep columns needed for sequence analysis
  select(seq_ID,
         event_hilo_BB) |>
 # create time index variable per sequence
  mutate(time_idx = seq(1:n())) 

#### TOL63
# set up df for sequence analysis  
selns_seq_tol63 <- hilo_cat |>
  # filter out off effort for now
  filter(seq_ID != "Off_eff_transit") |>
  ungroup()|>
  group_by(seq_ID) |>
  # only keep columns needed for sequence analysis
  select(seq_ID,
         event_hilo_tol63) |>
 # create time index variable per sequence
  mutate(time_idx = seq(1:n())) 


#### TOL125
# set up df for sequence analysis  
selns_seq_tol125 <- hilo_cat |>
  # filter out off effort for now
  filter(seq_ID != "Off_eff_transit") |>
  ungroup()|>
  group_by(seq_ID) |>
  # only keep columns needed for sequence analysis
  select(seq_ID,
         event_hilo_tol125) |>
 # create time index variable per sequence
  mutate(time_idx = seq(1:n())) 


#### TOL2000
# set up df for sequence analysis  
selns_seq_tol2k <- hilo_cat |>
  # filter out off effort for now
  filter(seq_ID != "Off_eff_transit") |>
  ungroup()|>
  group_by(seq_ID) |>
  # only keep columns needed for sequence analysis
  select(seq_ID,
         event_hilo_tol2k) |>
 # create time index variable per sequence
  mutate(time_idx = seq(1:n())) 




# create key for sequence IDs to fishing vs. transit
fish_states <- selns_seq |>
  select(seq_ID) |>
  mutate(fish_state = case_when(grepl("fishing", seq_ID)==TRUE ~ "fishing",
                   TRUE ~ "transit"))

```

### Descriptives

How many sequences?
How many events per sequence?
How many events of each type across the dataset?

```{r describe dataset}

length(unique(selns_seq$seq_ID))

length(unique(fish_states$seq_ID))

```

 

```{r events}

# descriptives for N events per sequence (same for all metrics)
event_summ <- selns_seq_BB |>
  group_by(seq_ID) |>
  summarize(n_events = n()) |>
  summarize(mean_n = mean(n_events),
            med_n = median(n_events),
            min_n = min(n_events),
            max_n = max(n_events))
event_summ

#### BB SPL
behav_table_BB <- table(selns_seq_BB$event_hilo_BB)

behav_table_BB


#### TOL63

behav_table_tol63 <- table(selns_seq_tol63$event_hilo_tol63)

behav_table_tol63

#### TOL125

behav_table_tol125 <- table(selns_seq_tol125$event_hilo_tol125)

behav_table_tol125

#### TOL2000

behav_table_tol2k <- table(selns_seq_tol2k$event_hilo_tol2k)

behav_table_tol2k




```

### Creating sequences

```{r reformat}

# reformat from long to wide

#### BB SPL
selns_seq_wide_BB <- selns_seq_BB |>
  filter(seq_ID != "pre_survey_Off_eff_transit",
         seq_ID != "post_survey_Off_eff_transit") |>
  pivot_wider(names_from = time_idx,
              values_from = event_hilo_BB)


#### TOL63
selns_seq_wide_tol63 <- selns_seq_tol63 |>
  filter(seq_ID != "pre_survey_Off_eff_transit",
         seq_ID != "post_survey_Off_eff_transit") |>
  pivot_wider(names_from = time_idx,
              values_from = event_hilo_tol63)

#### TOL125
selns_seq_wide_tol125 <- selns_seq_tol125 |>
  filter(seq_ID != "pre_survey_Off_eff_transit",
         seq_ID != "post_survey_Off_eff_transit") |>
  pivot_wider(names_from = time_idx,
              values_from = event_hilo_tol125)

#### TOL2000
selns_seq_wide_tol2k <- selns_seq_tol2k |>
  filter(seq_ID != "pre_survey_Off_eff_transit",
         seq_ID != "post_survey_Off_eff_transit") |>
  pivot_wider(names_from = time_idx,
              values_from = event_hilo_tol2k)





```

Create alphabet and assign labels

```{r alphabet}

# same for all metrics

# this contains categories that appear in the dataset
event_alphabet <- c("low_long",  "high_long", "low_short",  "high_short")

# formally assigning labels which might be helpful later
event_labels <- c("low_long",  "high_long", "low_short",  "high_short")

# assign colors (this feels unnecessary but sure)
event_colors <- viridis::plasma(n = 4, begin = 0.25, end = 0.75) 

```

Create sequence object 

```{r sequence}

# define sequence using seqdef from TraMineR package
event_seq_BB <- seqdef(selns_seq_wide_BB, 
                    var = 2:15,
                    alphabet = event_alphabet,
                    labels = event_labels,
                    xtstep = 2,
                    cpal = event_colors)


event_seq_tol63 <- seqdef(selns_seq_wide_tol63, 
                    var = 2:15,
                    alphabet = event_alphabet,
                    labels = event_labels,
                    xtstep = 2,
                    cpal = event_colors)


event_seq_tol125 <- seqdef(selns_seq_wide_tol125, 
                    var = 2:15,
                    alphabet = event_alphabet,
                    labels = event_labels,
                    xtstep = 2,
                    cpal = event_colors)

event_seq_tol2k <- seqdef(selns_seq_wide_tol2k, 
                    var = 2:15,
                    alphabet = event_alphabet,
                    labels = event_labels,
                    xtstep = 2,
                    cpal = event_colors)




# plot sequences

seqIplot(event_seq_BB, main = "BB SPL")
seqIplot(event_seq_tol63, main = "TOL63")
seqIplot(event_seq_tol125, main = "TOL125")
seqIplot(event_seq_tol2k, main = "TOL2k")

```


Cost matrix

- Use standard cost matrix values:
insertion/deletion cost = 1.0
missingness cost = 0.5*[highest cost in matrix]
substitution cost = constant


```{r cost and dist}

# same for all metrics
costmatrix <- seqsubm(event_seq, 
                      method="CONSTANT",
                      cval = 2, 
                      with.missing=FALSE,
                      miss.cost=1, 
                      time.varying=FALSE, 
                      weighted=TRUE,
                      transition="both", 
                      lag=1)

costmatrix


# optimal matching dissimilarity matrix

dist_om_BB <- seqdist(event_seq_BB, 
                   method = "OM", 
                   indel = 1.0, # insertion/deletion cost = 1
                   sm = costmatrix, # sub matrix = cost matrix
                   with.missing = FALSE)

dist_om_tol63 <- seqdist(event_seq_tol63, 
                   method = "OM", 
                   indel = 1.0, # insertion/deletion cost = 1
                   sm = costmatrix, # sub matrix = cost matrix
                   with.missing = FALSE)

dist_om_tol125 <- seqdist(event_seq_tol125, 
                   method = "OM", 
                   indel = 1.0, # insertion/deletion cost = 1
                   sm = costmatrix, # sub matrix = cost matrix
                   with.missing = FALSE)

dist_om_tol2k <- seqdist(event_seq_tol2k, 
                   method = "OM", 
                   indel = 1.0, # insertion/deletion cost = 1
                   sm = costmatrix, # sub matrix = cost matrix
                   with.missing = FALSE)

####################################################
##### STOPPED HERE WITH MULTIPLE METRICS EDITS #####
####################################################

```


### Clustering

```{r cluster}

event_clusters <- agnes(dist_om, diss = TRUE, method = "ward")
plot(event_clusters, which.plot = 2)


# cut into 2 clusters
event_clust_cut <- cutree(event_clusters, k = 2) |> 
  factor(labels = paste("Cluster", 1:2))

# plot sequences
seqplot(event_seq, 
        group = event_clust_cut, type = "I")



```


### Group differences between clusters

Visualize dendrogram with cluster ID
Run logistic regression to test whether vessel behavior  1/0 (fishing/transit) differs by cluster membership

```{r test clusters}

# Test cluster stability with bootstrapping

cluster_boot <- clusterboot(dist_om, 
            clustermethod = hclustCBI, 
            method = "ward.D2",
            k = 2,
            seed = 13958)

# get Jaccard coefficient for each cluster
# hoping for > 0.85 = 'highly stable' or at least > 0.6 = 'likely to reflect a pattern'
cluster_boot$bootmean
cluster_boot$bootbrd

# does vessel behavior differ by cluster membership?
# cluster ID = predictor
# vessel behav = response, coded as 1/0 (fishing/transit)

# add cluster labels to seq ID to associate w/fishing vs. transit
seq_clust_lab <- selns_seq_wide |>
  select(seq_ID) |>
  add_column(clust_ID = event_clust_cut)|>
  mutate(fish_state_binom = case_when(grepl("fishing", seq_ID)==TRUE ~ 1,
                   TRUE ~ 0),
         leaf_label = paste(seq_ID, fish_state_binom))

event_clusters$order.lab <- seq_clust_lab$leaf_label[event_clusters$order]
clust_colors = viridis::mako(n = 2, end = 0.75)

fviz_dend(event_clusters, 
          k = 2,
          color_labels_by_k = TRUE,
          k_colors = clust_colors,
          labels_font = "bold",
        
          rect = TRUE,
          rect_border = clust_colors,
          rect_fill = TRUE,
          
          lwd=1,
          horiz = TRUE,
          ggtheme = theme_bw()) +
  theme(axis.text = element_text(size = 10),
        axis.title.x = element_text(size = 12),
        title = element_blank())
ggsave("figs/dendrogram_k2.png", height=8, width=10, units = "in", dpi = 300)

```

Some observations from initial dendrogram:

  - fishing from lines 1 and 2a clustered with off-effort maneuvering from those same lines but fishing lines 3, 4, and 2b were separated out (distance effect?)


Fisher's Exact Test to see if cluster membership associated with fishing
```{r fisher}
# Fisher's Exact test to see if cluster assignment is related to fishing activity

fisher_tab <- table(seq_clust_lab |> ungroup() |> select(clust_ID, fish_state_binom))
fisher.test(fisher_tab)

```

That's promising!
